Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-11 vertical-thermal-review.Rmd
finished revising thermal tolerances section

<!--short par giving background on thermal tolerance-->
The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. 
At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress]. 
Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021]. 
Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->.
Across species, sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020].
Thermal tolerances globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
For example, *Quercus muehlenbergii* Engelm. growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx. growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].

Based on their variation with the temperatures experienced by leaves, thermal tolerance metrics are hypothesized to be higher in sun-exposed upper canopy leaves than in understory leaves. 
In an Australian savanna, crowns of *Acacia papyrocarpa* Benth. showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. 
However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019]. 
The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

In Australia, along a vertical gradient within a very open canopy, Acacia papyrocarpa showed greater T50 and higher composite climate stress in the lower, sun-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and Tair, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants (Curtis et al., 2019). 
These patterns of higher heat tolerance in the warmest leaves are consistent with interspecific patterns across temperature gradients (O’Sullivan et al. 2018; Feeley et al., 2020; Slot et al., 2021a), and experienced maximum temperatures (Perez & Feeley, 2020). 
However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences.

>>>>

<!--short par giving background on thermal tolerance-->
The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. 
At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress]. 
Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021]. 
Thermal tolerances vary across species, being more closely linked to leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->, and with higher thermal tolerance among leaves that experience higher maximum temperatures [@perezPhotosyntheticHeatTolerances2020].
$T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
<!--For example, *Quercus muehlenbergii* Engelm. growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx. growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. (removed because I had trouble finding this in the paper, plus it's not really needed, but feel free to reinstate) -->

<!--par on thermal tolerance across vertical gradients-->
Based on their variation with the temperatures experienced by leaves, thermal tolerance metrics are hypothesized to be higher in sun-exposed upper canopy leaves than in understory leaves. 
Indeed, along a vertical gradient within a very open canopy in Australia, crowns of *Acacia papyrocarpa* Benth. showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. 
However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019]. 
This small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences.
